---
title: "Capstone Final Project"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

**Michelle Turovsky**
<br>
**mturo2**

***

<br>

##Section 1 - Dataset Overview

####Setup environment
```{r, warning = FALSE, message = FALSE}
options(scipen=999)
options(warn=-1)
library(tidyr)
library(dplyr)
library(caret)
library(glmnet)
library(stringr)
library(datapasta)
library(fuzzyjoin)
library(data.table)
library(ggplot2)
library(ggpubr)
library(FactoMineR)
library(factoextra)
library(rpart.plot)
library(klaR)
library(keras)
library(tensorflow)
```

<br>

####Download and view metadata
```{r}
outcomes <- read.csv("~/train.csv", sep=",", stringsAsFactors = FALSE, header = TRUE, na.strings="", encoding="UTF-8")

str(outcomes)
summary(outcomes)
summary(is.na(outcomes))


colnames(outcomes)
length(unique(outcomes$Name))
length(unique(outcomes$DateTime))
length(unique(outcomes$OutcomeType))
length(unique(outcomes$AnimalType))
length(unique(outcomes$SexuponOutcome))
length(unique(outcomes$AgeuponOutcome))
length(unique(outcomes$Breed))
length(unique(outcomes$Color))

table(outcomes$OutcomeType)
```

<br>

###Section 2 - PreProcessing Data

####Feature Removal
```{r}
outcomes<-outcomes %>% 
  dplyr::select(-AnimalID, -OutcomeSubtype)
```

####Missing Data Removal
```{r}
outcomes<-outcomes %>% 
  filter(!is.na(SexuponOutcome)) %>% 
  filter(!is.na(AgeuponOutcome))
```

####Evaluate Feature Significance
```{r}
###Look at outcome variable
outcomes %>% 
  group_by(OutcomeType) %>% 
  summarize(n=n()) %>% 
  ggplot(., aes(x=OutcomeType, y=n))+
  geom_bar(stat="identity")

###Evaluate Feature Significance 
tbl1 <- with(outcomes, table(SexuponOutcome, OutcomeType))
barplot(tbl1, beside = TRUE, legend = TRUE)
chisq.test(tbl1)

tbl2 <- with(outcomes, table(AnimalType, OutcomeType))
barplot(tbl2, beside = TRUE, legend = TRUE)
chisq.test(tbl2)

tbl3 <- with(outcomes, table(OutcomeType, AgeuponOutcome))
barplot(tbl3, beside = TRUE, legend = TRUE)
chisq.test(tbl3)
```

<br>

####Feature Engineering
```{r}
######################################################
#################  Name Variable  ####################
######################################################
outcomes<-outcomes %>% 
  mutate(HasName = ifelse(is.na(Name), 0 ,1)) %>% 
  dplyr::select(-Name)

###Evaluate Feature Significance
tbl4 <- with(outcomes, table(HasName, OutcomeType))
barplot(tbl4, beside = TRUE, legend = TRUE)
chisq.test(tbl4)
```

<br>

```{r}
######################################################
################ DateTime Variable ###################
######################################################
outcomes<-outcomes %>% 
  mutate(Month=as.numeric(substr(DateTime, 6, 7))) %>% 
  mutate(Hour=as.numeric(substr(DateTime, 12, 13))) %>% 
  dplyr::select(-DateTime)

###Evaluate Feature Significance
tbl5 <- with(outcomes, table(Month, OutcomeType))
barplot(tbl5, beside = TRUE, legend = TRUE)
chisq.test(tbl5)

tbl6 <- with(outcomes, table(OutcomeType, Hour))
barplot(tbl6, beside = TRUE, legend = TRUE)
chisq.test(tbl6)

outcomes<-outcomes %>%
  mutate(Month_Group=ifelse(Month %in% c(1,2,3), "Q1",
                      ifelse(Month %in% c(4,5,6), "Q2",
                             ifelse(Month %in% c(7,8,9), "Q3", "Q4")))) %>%
  dplyr::select(-Month)

###Evaluate Feature Significance
tbl7 <- with(outcomes, table(Month_Group, OutcomeType))
barplot(tbl7, beside = TRUE, legend = TRUE)
chisq.test(tbl7)

outcomes<-outcomes %>% 
  mutate(Hour_Group=ifelse(Hour<6, "Night",
                     ifelse(Hour>=6 & Hour <12, "Morning",
                            ifelse(Hour>=12 & Hour <17, "Afternoon",
                                   ifelse(Hour>=17 & Hour <19, "Evening", "Night")))))%>% 
  dplyr::select(-Hour)

###Evaluate Feature Significance
tbl8 <- with(outcomes, table(Hour_Group, OutcomeType))
barplot(tbl8, beside = TRUE, legend = TRUE)
chisq.test(tbl8)
res.ca <- CA(tbl8, graph = FALSE)
fviz_ca_biplot(res.ca, repel = TRUE)

```

<br>

```{r}
######################################################
############# AgeuponOutcome Variable ################
######################################################
outcomes<-outcomes %>% 
  separate(AgeuponOutcome, c("Age_Number", "Age_Type"))

unique(outcomes$Age_Type)

outcomes<-outcomes %>% 
  mutate(Age_Number=as.numeric(Age_Number)) %>% 
  mutate(AgeuponOutcome=ifelse(Age_Type %in% c("year", "years"), ceiling(Age_Number*365), Age_Number)) %>% 
  mutate(AgeuponOutcome=ifelse(Age_Type %in% c("month", "months"), ceiling(Age_Number*30), AgeuponOutcome)) %>%
  mutate(AgeuponOutcome=ifelse(Age_Type %in% c("week", "weeks"), ceiling(Age_Number*7), AgeuponOutcome))

age_box<-boxplot(as.numeric(outcomes$AgeuponOutcome)~outcomes$OutcomeType,
     col="lightblue",
     xlab="OutcomeType",
     ylab="Age (Days)")

age_box$names
age_box$stats

outcomes<-outcomes %>%
  dplyr::select(-Age_Number, -Age_Type) %>%
  mutate(AgeuponOutcome=ifelse(AgeuponOutcome<=30, "OneMonth",
                                      ifelse(AgeuponOutcome<=60, "TwoMonth",
                                              ifelse(AgeuponOutcome>60 & AgeuponOutcome<=180, "HalfYear",
                                             ifelse(AgeuponOutcome>180 & AgeuponOutcome<=360, "Year",
                                                    ifelse(AgeuponOutcome>360 & AgeuponOutcome<=750, "TwoYear", 
                                                           ifelse(AgeuponOutcome>750 & AgeuponOutcome<=2190, "Adult", "OldAdult")))))))

###Evaluate Feature Significance
tbl9 <- with(outcomes, table(AgeuponOutcome, OutcomeType))
barplot(tbl9, beside = TRUE, legend = TRUE)
chisq.test(tbl9)

ggballoonplot(as.data.frame(tbl9), fill = "value")+
  scale_fill_viridis_c(option = "C")

res.ca <- CA(tbl9, graph = FALSE)
fviz_ca_biplot(res.ca, repel = TRUE)
```

<br>

```{r}
######################################################
############# SexuponOutcome Variable ################
######################################################
outcomes<-outcomes %>% 
  separate(SexuponOutcome, c("Neuter", "Gender")) 

outcomes<-outcomes %>% 
  mutate(Gender=ifelse(is.na(Gender), "Unknown", Gender)) %>% 
  mutate(Neuter=ifelse(Neuter=="Spayed", "Neutered", Neuter))

###Evaluate Feature Significance
tbl10 <- with(outcomes, table(Neuter, OutcomeType))
barplot(tbl10, beside = TRUE, legend = TRUE)
chisq.test(tbl10)

ggballoonplot(as.data.frame(tbl10), fill = "value")+
  scale_fill_viridis_c(option = "C")

tbl11 <- with(outcomes, table(Gender, OutcomeType))
barplot(tbl11, beside = TRUE, legend = TRUE)
chisq.test(tbl11)

ggballoonplot(as.data.frame(tbl11), fill = "value")+
  scale_fill_viridis_c(option = "C")

res.ca <- CA(tbl11, graph = FALSE)
fviz_ca_biplot(res.ca, repel = TRUE)

```

<br>

```{r}
######################################################
################# Color Variable #####################
######################################################

outcomes2<-outcomes %>% 
  mutate(Tabby_Tiger=ifelse(str_detect(Color, "Tabby"), 1, 0)) %>% 
  mutate(Tabby_Tiger=ifelse(str_detect(Color, "Tiger"), 1, Tabby_Tiger)) %>% 
  mutate(Brindle=ifelse(str_detect(Color, "Brindle"), 1, 0)) %>% 
  mutate(Merle=ifelse(str_detect(Color, "Merle"), 1, 0)) %>%
  mutate(Tick=ifelse(str_detect(Color, "Tick"), 1, 0)) %>% 
  mutate(Point=ifelse(str_detect(Color, "Point"), 1, 0)) 


outcomes3<-outcomes2 %>% 
    mutate(Color=gsub("Tabby", "", Color)) %>% 
    mutate(Color=gsub("Tiger", "", Color)) %>% 
    mutate(Color=gsub("Brindle", "", Color)) %>% 
    mutate(Color=gsub("Merle", "", Color)) %>% 
    mutate(Color=gsub("Tick", "", Color)) %>% 
    mutate(Color=gsub("Point", "", Color)) %>% 
    mutate(Color=gsub("/", " ", Color)) %>% 
    mutate(Color=gsub("\\s+"," ", Color)) %>% 
    separate(Color, c("Color1", "Color2"), sep=" ") %>%
    dplyr::select(-Color2) %>% 
    rename(Color=Color1)

length(unique(outcomes3$Color))
table(outcomes3$Color)

###Evaluate Feature Significance
tbl12 <- with(outcomes3, table(Color, OutcomeType))
chisq.test(tbl12)

ggballoonplot(as.data.frame(tbl12), fill = "value")+
  scale_fill_viridis_c(option = "C")

```

<br>

```{r}
######################################################
################# Breed Variable #####################
######################################################
dog_groups <- read.csv("~/dog_groups.csv", sep=",", stringsAsFactors = FALSE, header = TRUE, na.strings="", encoding="UTF-8") %>% 
  dplyr::select(-X)

length(unique(outcomes3$Breed))

outcomes4<-outcomes3 %>% 
  mutate(Mix=ifelse(str_detect(Breed, "Mix"), 1, 0)) %>% 
  mutate(Mix=ifelse(str_detect(Breed, '/'), 1, Mix)) %>% 
  mutate(Breed=gsub("Mix", "", Breed))

head(unique(outcomes4$Breed),10)

outcomes5<-outcomes4 %>% 
  separate(Breed, c("Breed1", "Breed2"), sep="/") %>% 
  mutate(Breed2=ifelse(is.na(Breed2), "", Breed2)) %>% 
  mutate(Breed1=trimws(Breed1)) %>% 
  mutate(Breed2=trimws(Breed2)) %>% 
  mutate(Breed1=gsub("\\bTerr\\b", "Terrier", Breed1)) %>% 
  mutate(Breed2=gsub("\\bTerr\\b", "Terrier", Breed2))  
  
grp1<-left_join(outcomes5, dog_groups, by=c("Breed1"="Breed"))
grp2<-left_join(grp1, dog_groups, by=c("Breed2"="Breed"))
grp3<-grp2 %>% 
  mutate(Group.x=ifelse(is.na(Group.x), Group.y, Group.x)) %>% 
  mutate(Group.x=ifelse(is.na(Group.x), "Unknown", Group.x)) %>% 
  mutate(Group.y=ifelse(Group.y==Group.x,NA, Group.y)) %>% 
  mutate(Final_Group=paste0(Group.x, "," ,Group.y)) 
grp3$Final_Group<-gsub(",NA", "", grp3$Final_Group)

grp3$Final_Group<-sapply(strsplit(as.character(grp3$Final_Group), ","), function(x) paste(sort(x), collapse=" "))
  
outcomes6<-grp3 %>% 
  mutate(Final_Group=ifelse(AnimalType=="Cat", Breed1, Final_Group)) %>% 
  dplyr::select(-Breed1, -Breed2, -Group.x, -Group.y) 

#Cat groups with less than 50 instances
cat_grps<-outcomes6 %>% 
  filter(AnimalType=="Cat") %>% 
  group_by(Final_Group) %>% 
  summarize(n=n()) %>% 
  filter(n<50)

outcomes6<-outcomes6 %>% 
  mutate(Final_Group=ifelse(Final_Group %in% cat_grps$Final_Group, "Other", Final_Group))
  
length(unique(outcomes6$Final_Group))

###Evaluate Feature Significance
tbl13 <- with(outcomes6, table(Final_Group, OutcomeType))
chisq.test(tbl13)
```

<br>

###Section 3 - Modeling

####One Hot Encode
```{r}
outcomes6[,2:ncol(outcomes6)] <- lapply(outcomes6[,2:ncol(outcomes6)], factor)

###Make dummy variables - full rank matrix
dmy <- dummyVars("OutcomeType ~ .", data = outcomes6, fullRank=T)
trsf<- data.frame(predict(dmy, newdata = outcomes6))
outcomes7<-cbind(outcomes6$OutcomeType, trsf) %>% 
  rename(OutcomeType='outcomes6$OutcomeType')
head(trsf)

outcomes7[,2:ncol(outcomes7)] <- lapply(outcomes7[,2:ncol(outcomes7)], factor)

```

####Train Test Split

When creating my final train and test datasets I initially attempted to combat my imbalanced dataset by upsampling my minority classes, and creating sythentic data with SMOTE to see if model accuracy would improve. However after running various models on these datasets, I found accuracy to decrease significantly. Therefore I have left the following code here to document my work, but will not run it for the purposes of my final output.

<br>

Upsampling and SMOTE datasets
```{r}
########################################
############### SMOTE ##################
########################################
# library(DMwR)
# set.seed(1)
# in_train <- createDataPartition(outcomes7$OutcomeType, p = 0.8, list = FALSE)
# outcomes_train <- outcomes7[in_train, ]
# sm<-outcomes_train %>%
#   mutate(OutcomeType=as.factor(ifelse(OutcomeType=="Died", "Died", "common")))
# newData <- SMOTE(OutcomeType ~ ., sm, perc.over = 2500, perc.under=100)
# newData<-newData %>%
#   as.data.frame() %>%
#   filter(OutcomeType=="Died") %>%
#   droplevels()
# outcomes_train<-rbind(newData, outcomes_train) %>%
#   droplevels()
# outcomes_test <- outcomes7[-in_train, ]
 
########################################
############ Upsampling ################
########################################

# outcomes_samp<-as.data.frame(upSample(x=outcomes_train[,-1], y=as.factor(outcomes_train[,1]), list=TRUE))
# colnames(outcomes_samp)<-c(colnames(outcomes_train[,-1]), "OutcomeType")
# head(outcomes_samp)
# table(outcomes_samp$OutcomeType)
# outcomes_train<-outcomes_samp
# outcomes_test <- outcomes7[-in_train, ]
```

<br>

Final Train-Test Datasets
```{r}
set.seed(1)
in_train <- createDataPartition(outcomes7$OutcomeType, p = 0.8, list = FALSE)

outcomes_train <- outcomes7[in_train, ]
outcomes_test <- outcomes7[-in_train, ]
```

<br>

####RPART
```{r}

set.seed(1)
rp <- caret::train(
  as.factor(OutcomeType) ~., data = outcomes_train, 
          method="rpart",
  trControl = trainControl("cv", number = 5))

rp

predictions=predict(rp, outcomes_test)
confusionMatrix(predictions, outcomes_test$OutcomeType)

rpart.plot(rp$finalModel)

```

<br>

####Random Forest
```{r}
rf_grid <- expand.grid(mtry = c(5, 15, 25))

Sys.time()
set.seed(1)
rf <- caret::train(
  as.factor(OutcomeType) ~., data = outcomes_train, method = "rf",
  tuneGrid = rf_grid, 
  trControl = trainControl("cv", number = 5))
Sys.time()
rf

varImp(rf)

predictions_rf=predict(rf, outcomes_test)
confusionMatrix(predictions_rf, outcomes_test$OutcomeType)

```

<br>

####KNN 
```{r}
Sys.time()
set.seed(1)
knn <- caret::train(
  OutcomeType ~., data = outcomes_train, method = "knn",
  tuneGrid   = expand.grid(k = c(2,10,15)),
  trControl = trainControl("cv", number = 5))
Sys.time()
knn

predictions_knn=predict(knn, outcomes_test)
confusionMatrix(predictions_knn, outcomes_test$OutcomeType)
```

<br>

####Naive Bayes
```{r}
Sys.time()
set.seed(1)
nb <- caret::train(y=outcomes_train$OutcomeType,
                   x=outcomes_train[,-1], method = "nb",
                   trControl = trainControl("cv", number = 5))
Sys.time()
nb

predictions_nb=predict(nb, outcomes_test)
confusionMatrix(predictions_nb, outcomes_test$OutcomeType)
```

<br>

####GBM
```{r}
Sys.time()
set.seed(1)
gbm <- caret::train(
  OutcomeType ~., data = outcomes_train, method = "gbm",
  trControl = trainControl("cv", number = 5))
Sys.time()
gbm
summary(gbm)

predictions_gbm=predict(gbm, outcomes_test)
confusionMatrix(predictions_gbm, outcomes_test$OutcomeType)

```

<br>

####ANN - MLP
```{r}
outcomes_train <- outcomes7[in_train, ]
outcomes_test <- outcomes7[-in_train, ]
```

```{r}
train_labels<-as.data.frame(outcomes_train[,1 ])
colnames(train_labels)<-c("OutcomeType")
outcomes_train <- outcomes_train[,-1 ]

test_labels <- as.data.frame(outcomes_test[,1 ])
colnames(test_labels)<-c("OutcomeType")
outcomes_test <- outcomes_test[,-1 ]

dmy <- dummyVars(" ~ .", data = test_labels)
test_labels<- data.frame(predict(dmy, newdata = test_labels))

dmy <- dummyVars(" ~ .", data = train_labels)
train_labels<- data.frame(predict(dmy, newdata = train_labels))
```


```{r}
Sys.setenv(KERAS_IMPLEMENTATION='tensorflow')
library(keras)
model <- keras_model_sequential() %>%
  layer_dense(units = 5, activation = "relu",
              input_shape = dim(outcomes_train)[2]) %>%
  layer_dense(units = 5, activation = "softmax")

model %>% compile(
  loss = "categorical_crossentropy",
  optimizer = "adam",
  metrics = "accuracy")
model

set.seed(1)
history <- model %>% fit(as.matrix(outcomes_train), 
                         as.matrix(train_labels),
                         batch_size=50,
                         epochs = 10,
                      validation_data=list(as.matrix(outcomes_test),as.matrix(test_labels)))

plot(history)
```

```{r}
predictions=model %>% predict_classes(as.matrix(outcomes_test), batch_size = 128)
```

```{r}
plot(history$metrics$acc, main="Model Accuracy", xlab = "epoch", ylab="accuracy", col="blue", type="l")
# Plot the accuracy of the validation data
lines(history$metrics$val_acc, col="green")
# Add Legend
legend("bottomright", c("train","test"), col=c("blue", "green"), lty=c(1,1))
```

```{r}
outcomes_test <- outcomes7[-in_train, ]
labs<-outcomes_test[,1] 
predictions<-predictions %>%
  as.data.frame() %>% 
  mutate(preds=ifelse(.==0, "Adoption", ifelse(.==1, "Died",
                                        ifelse(.==2, "Euthanasia",
                                               ifelse(.==3, "Return_to_owner", "Transfer"))))) %>% 
  dplyr::select(preds) %>% 
  as.vector()
table(predictions$preds,as.vector(labs))
```

```{r}
score <- model %>% evaluate(as.matrix(outcomes_test[,-1]), as.matrix(test_labels), batch_size = 128)
print(score)
```